---
title: "Hydraulics: pipes and open channel"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
```

## Case study: Jostedal

Use NVE Atlas, Norgeskart, Nevina to find the catchments specific to each water intake in Jostedalen.

Calculate much flow can be achieved in the tunnel

How much flow can be directed from the side-catchments back to the main reservoir by gravity.
Will one of the side intake need a non-return valve to avoid a backflow through the intake?
Experiment with the behavior of those intakes during various events: intense hourly precipitation, big daily events...



In this tutorial, you will learn how to summarise a table of data, including:

* How to summarise tables with `summarise()` 
* How to recognize the most useful families of functions to combine with `summarise()`
* How to combine multiple dplyr operations with the pipe, `%>%`
* How to compute counts of observations with `n()`
* How to summarise groups of observations with `group_by()` and `summarise()`

The readings in this tutorial follow [_R for Data Science_](http://r4ds.had.co.nz/), section 5.6.

### Setup


I've preloaded the packages for this tutorial with 

```{r eval = FALSE}
library(tidyverse) # loads dplyr, ggplot2, and others
```
    

    