---
title: "Create new variables"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
```

## Hydrological modelling

In this exercise session focusing on the Songdal catchment, we will experiment with various hydrological models:

* Recession from a linear reservoir
* Response to a snowmelt event
* Hydrological extremes for Songdalselvi

The course material is [hydrological processes](./01_hydrology/01_hydrological_processes)

### Setup

Download meteorological and hydrological data for the Sogndalselvi catchment.

To visualize the data, we will use the ggplot2 package that you met in [Data Visualization Basics](../02-data-vis-basics/02-data-vis-basics.html). 

I've preloaded the packages for this tutorial with 

```{r eval = FALSE}
library(tidyverse) # loads dplyr, ggplot2, and others
```

## Add new variables with mutate()

A data set often contains information that you can use to compute new variables. `mutate()` helps you compute those variables. Since `mutate()` always adds new columns to the end of a dataset, we'll start by creating a narrow dataset which will let us see the new variables (If we added new variables to `flights`, the new columns would run off the side of your screen, which would make them hard to see). 


<div id="summariseex3-hint">
**Hint:** Don't forget to use `na.rm = TRUE` where appropriate. 
</div>

```{r summariseex3-check}
"Wow! You did awesome."
```

